Density-dependent selection and the limits of relative fitness

### 2017-12-22 manuscript.tex
l

In the previous section we showed that selection and the regulation of population density can be independent even if population growth is density-regulated. Nevertheless, selection and density regulation \textit{are} intimately connected in widely used models of population growth, as well as for the $b$ and $d$ traits in our variable-density lottery model.

To see why this potentially poses a threat to relative fitness, consider the simple birth-death model \cite[pp. 20]{kostitzin_1939} \citep{travis_2013} 
\begin{equation}
\frac{d n_i}{dt}=(b_i -\delta_iN) n_i \label{eq:simplebirthdeath}
\end{equation}
where $\delta_i$ is the per-capita increase in mortality rate due to crowding (for simplicity, there are no deaths when uncrowded), playing a similar role as $K$ in the logistic model.

@@

In practice the density dependence in Eq.~\eqref{eq:Ndependentsweep} only matters if $N$ changes substantially during a sweep. This can easily occur if a population is far from demographic equilibrium (we return to this scenario in the Discussion). However, even if $N$ has reached equilibrium, it will change substantially over a $\delta$-sweep if selection on $\delta$ is sufficiently strong. To quantify this effect, we need to account for how much $N$ changes as a result of a $\delta$-sweep beginning and ending in equilibrium \citep{kimura1969natural}; from Eq.~\eqref{eq:simplebirthdeath} we have an increase from $N_{\rm initial}=b_i/\delta_i$ to $N_{\rm final}=b_i/\delta_i(1-\epsilon)=N_{\rm initial}/(1-\epsilon)$. The corresponding selection coefficient increases from $s_{\rm initial}= \epsilon b_i$ to $s_{\rm final}=s_{\rm initial}/(1-\epsilon)$. Consequently, substantial deviations from Eq.~\eqref{eq:canonical} occur with proportional changes to $\delta$ of order $\epsilon=0.1$ and upwards (Fig.~\ref{fig:strengthofselection}).

\begin{figure}
\centering
\includegraphics[scale=0.8]{strengthofselection.pdf}
\caption{\label{fig:strengthofselection} Proportional change in the selection coefficient over a ``$K$-like'' sweep for a type that experiences proportionally $1-\epsilon$ fewer deaths induced by crowding. The population is in demographic equilibrium at the start and end of the sweep.}
\end{figure}

Let us now turn to selection on $b$ and $d$ in our lottery model. Recall that $m_i=b_i n_i U/T$, and so $L=M/U=\overline{b}N/T$ where $\overline{b}$ is the population mean $b$. Thus, from Eq.~\eqref{eq:masterequalc} we have
\begin{equation}
\Delta n_i = \left(\frac{b_i}{\overline{b}}\frac{1-e^{-\overline{b}N/T}}{N}(T-N)-d_i\right)n_i. \label{eq:bdensitydependence}
\end{equation}
It can be seen that the mortality $d$ is akin to the birth rate in Eq.~\eqref{eq:simplebirthdeath}, and so, while $d$ does affect density, selection on $d$ is density independent. Thus, $d$ sweeps follow the canonical relative fitness model exactly.

@@

>>>>

In the previous section we showed that $c$-selection and the regulation of population density are independent even though population growth is density-regulated in our variable-density lottery. Nevertheless, selection and density regulation \textit{are} intimately connected in widely used models of population growth, as well as for the lottery $b$ and $d$ traits.

To see why this connection potentially poses a threat to relative fitness, consider the simple birth-death model \cite[pp. 20]{kostitzin_1939} \citep{travis_2013} 
\begin{equation}
\frac{d n_i}{dt}=(b_i -\delta_iN) n_i \label{eq:simplebirthdeath}
\end{equation}
where $\delta_i$ is the per-capita increase in mortality rate due to crowding (for simplicity, there are no deaths when uncrowded), playing a similar role as $K$ in the logistic model.

@@

In practice the density dependence in Eq.~\eqref{eq:Ndependentsweep} only matters if $N$ changes substantially during a sweep. This can easily occur if a population is far from demographic equilibrium (we return to this scenario in the Discussion). However, even if $N$ has reached equilibrium, it will change substantially over a $\delta$-sweep if selection on $\delta$ is sufficiently strong. To quantify this effect, we need to account for how much $N$ changes as a result of a $\delta$-sweep beginning and ending in equilibrium \citep{kimura1969natural}; from Eq.~\eqref{eq:simplebirthdeath} we have an increase from $N_{\rm initial}=b_i/\delta_i$ to $N_{\rm final}=b_i/\delta_i(1-\epsilon)=N_{\rm initial}/(1-\epsilon)$. The corresponding selection coefficient increases from $s_{\rm initial}= \epsilon b_i$ to $s_{\rm final}=s_{\rm initial}/(1-\epsilon)$. Consequently, noticeable deviations from Eq.~\eqref{eq:canonical} occur with proportional changes to $\delta$ of order $\epsilon=0.2$ and upwards i.e. selection must be quite strong (Fig.~\ref{fig:strengthofselection}).

\begin{figure}
\centering
\includegraphics[scale=0.8]{strengthofselection.pdf}
\caption{\label{fig:strengthofselection} (a) Proportional change in the selection coefficient over a ``$K$-like'' sweep for a type that experiences proportionally $1-\epsilon$ fewer deaths induced by crowding. The population is in demographic equilibrium at the start and end of the sweep. (b) Example equilibrium-to-equilibrium $\delta$-sweep (Eq.~\ref{eq:Ndependentsweep}) for $\epsilon=0.2$ showing a noticeable deviation from the canonical selection equation.}
\end{figure}

Let us now turn to selection on $b$ and $d$ in our lottery model. Recall that $m_i=b_i n_i U/T$, and so $L=M/U=\overline{b}N/T$ where $\overline{b}$ is the population mean $b$. Thus, from Eq.~\eqref{eq:masterequalc} we have
\begin{equation}
\Delta n_i = \left(\frac{b_i}{\overline{b}}\frac{1-e^{-\overline{b}N/T}}{N}(T-N)-d_i\right)n_i. \label{eq:bdensitydependence}
\end{equation}
It can be seen that the mortality $d$ is akin to the birth rate in Eq.~\eqref{eq:simplebirthdeath}, and so, while $d$ does affect density, selection on $d$ is density independent. Thus, $d$ sweeps follow the canonical relative fitness model exactly (Fig.~\ref{fig:DDS_lottery}).

@@

\begin{figure}
\centering
\includegraphics[scale=0.8]{DDS_lottery.pdf}
\caption{\label{fig:DDS_lottery} The density-dependence of selection in our variable-density lottery plotted as the selection coefficient $(\Delta n_j-\Delta n_i)/n_j$ experienced by an adaptive variant $j$ present at the same frequency as a wildtype $i$. Here $b_i=1$, $d_i=0.5$ and $c_i=1$. For $b$-selection we set $b_j=b_i(1+\epsilon)$, and similarly for $c$ and $d$, with $\epsilon=0.1$. $d$-selection is density-independent, $b$-selection gets weaker with lower territorial availability, while $c$-selection initially increases with density as territorial contests become more important, but eventually also declines as  available territories become scarce. The equilibrium density for the $i$ type is $\approx 0.4$.}
\end{figure}
